---
title: "Mini Data-Analysis Deliverable 1"
output: html_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1. Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

```
install.packages("devtools")
devtools::install_github("UBC-MDS/datateachr")
```

2. Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3. Make a repository in the https://github.com/stat545ubc-2021 Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Vincenzo as collaborators. 

4. When you go to submit, submit a URL to your repository to canvas. 

# Instructions

More details regarding the instructions and points allocated to each task can be found below. Briefly,

+ Each milestone is worth 30 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

+ 10 points will be allocated to the reproducibility, cleanliness, and coherence of the analysis. While the three milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as 3 chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with your dataset of choosing
+ Think of 4 questions that you would like to answer with your data
+ Generate a reproducible and clear report using R Markdown

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

+ *apt_buildings*: Acquired courtesy of The City of Toronto’s Open Data Portal. It currently has 3455 rows and 37 columns.

+ *building_permits*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 20680 rows and 14 columns.

+ *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

+ *flow_sample*: Acquired courtesy of The Government of Canada’s Historical Hydrometric Database. It currently has 218 rows and 7 columns.

+ *parking_meters*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 10032 rows and 22 columns.

+ *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

+ *vancouver_trees*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

+ We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

+ Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

1: *building_permits*    
2: *cancer_sample*    
3: *flow_sample*   
4: *vancouver_trees*   

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*. 

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

```{r}
#Exploring building_permits dataset
class(building_permits)
dim(building_permits)
glimpse(building_permits)

#Exploring flow_sample dataset
class(flow_sample)
dim(flow_sample)
glimpse(flow_sample)

#Exploring cancer_sample dataset
class(cancer_sample)
dim(cancer_sample)
glimpse(cancer_sample)

#Exploring vancouver_trees dataset
class(vancouver_trees)
dim(vancouver_trees)
glimpse(vancouver_trees)
```

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation. 

ANSWER: **I would be most interested to explore the flow_sample and vancouver_trees datasets. After using a glimpse() function to look more closely at the datasets, I could conclude that both of these dataset contain data related to the environment - i.e. data related to the different species of trees and the data on stream/river flows. These are the topics I am most interested in personally and professionally, and where I would be able to theoretically explain the correlations between variables if I find any.**

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

ANSWER: **Potential research question related to the vancouver_trees dataset: How did the city's preferences regarding what trees to plant have changed over time?**

**Potential research questions related to the flow_sample dataset: What are the patterns of annual variability of the streamflow?**

**I would prefer to explore flow_sample dataset**


# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

* [x] Plot the distribution of a numeric variable.
* [ ] Create a new variable based on other variables in your data (only if it makes sense)
* [x] Investigate how many missing values there are per variable. Can you find a way to plot this?
* [x] Explore the relationship between 2 variables in a plot.
* [x] Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
* [x] Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
* [x] Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
* [x] Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->

## Exploring Data and Distribution

First, the distribution of data should be explored to gain more understanding on what data the variables store. This would allow to determine, for example, whether the flow variable only stores maximum and minimum extreme events or other flow data.

To explore the data, the dataset shall firstly be converted to a tibble:

```{r}
flow_sample <- as_tibble(flow_sample)
class(flow_sample)
```

Next, **the distribution** of the flow will be explored. The following code produces histogram, which shows the frequency of different values of flow. The histogram has 30 bins and the missing values are omitted. 

```{r}
flow_distribution1 = ggplot(flow_sample, aes(flow)) +geom_histogram(bins=30,na.rm=TRUE)
print(flow_distribution1)
```

As can be seen from the histogram, the most common values taken by the variable "flow" are near-zero values.

Another way to visualize the distribution of flow values is by **plotting its density**:

```{r}
flow_distribution2 <- ggplot(flow_sample, aes(x = flow))+ geom_density(na.rm=TRUE)
print(flow_distribution2)
```

As can be seen from the density plot, the majority of values of flow are low - i.e. below 75, however there is an increase in the density around 200, which suggests that there is a considerable amount of values of flow around 200.

Based on the observed distribution of the flow data, as well as the fact that data is split in two categories ("maximum" and "minimum") by the categorical variable "extreme_type", the conclusion may be reached that the dataset only displays the extreme flow events.


This assumption can be verified by creating a boxplot. The boxplot with a logarithmic scale is plotted to ensure a meaningful visual representation considering a significant spread of values of flow. 

```{r}
flow_boxplot <- ggplot(flow_sample, aes(extreme_type, flow))+ geom_boxplot(width = 0.2)+scale_y_log10()
print(flow_boxplot)
```

The boxplot verified the assumption that the maximum and minimum categories correspond to maximum and minimum extreme flow events.


## Exploring Missing Values

Next, it is also important to explore the missing values in the dataset to understand the limitations of the data and hence develop research questions that can realistically be answered with this dataset.

To explore the missing values, the data frame will be created containing the number of missing values for each of the variables

```{r}
#creating a data frame containing names of the variables in the first column and the corresponding number of missing values in the second column
flow_missing <-data.frame(names(flow_sample), colSums(is.na(flow_sample)))

#Adding meaninful column names:
names(flow_missing)[1] <- "variable"
names(flow_missing)[2] <- "missing_val"
```

Bar chart can be plotted to represent the number missing values per variable:

```{r}
flow_missing_bar = ggplot(flow_missing, aes(variable,missing_val)) + geom_bar(stat="identity")
print(flow_missing_bar)
```


As can be seen from the plot, the majority of the missing values are in column "sym". Because of the significant differences in the number of missing data between "sym" and other variables, it is hard to visually assess how much data is missing in other columns. Hence, it may be valuable to look at the plot at logarithmic scale:

```{r}
flow_missing_bar +scale_y_log10()
```


From this plot it can be derived that few data is missing from columns "day", "flow", and "month", while large amont of data is missing for variable "sym"


## Exploring the relation between variables
Next, it would be valuable to explore the relation between month and extreme flow event. Particularly I would like to find during which months the maximum flow usually occurs, and during which months the minimum flow occurs.

### Creating new tibble holding a subset of flow_sample data

To achieve this, first, **a tibble should be created that only holds the variables "flow", "month", and "extreme_type" and does not have any missing data**. Missing data will be removed after creating a subset to preserve as much data as possible. For example, if missing data were removed before creating a subset, it may result in a loss of entire rows of data where only "day" or "sym" variables are missing. This would limit the analysis of the relation between month and flow.

```{r}
#Creating a subset of the data with columns "month", "flow", and "extreme_type"
flow_subset <- flow_sample %>%
    select(month, flow, extreme_type) %>%
    na.omit() #removing rows with missing values

print(flow_subset)
```


Next, it should be verified whether the "month" column is categorical, and if not its class should be changed accordingly

```{r}
class(flow_subset$month) 

flow_subset$month=as.factor(flow_subset$month) #this formula makes makes the column categorical

#Checking the class again:
class(flow_subset$month)
```


### Exploring the relation between variables "month" and "flow" using bar chart

To explore how the extreme flow events are distributed across the months, a bar chart will be created. While a bar char cannot provide an insight on the magnitude of the individual flow events that have occurred during each of the months, it provides a valuable information on the number of extreme flow events that have occurred during each of the months.

The following code plots the bar graph showing the number of the extreme flow events that have occurred each month. The colors differentiate between maximum and minimum flow events:

```{r}
flow_subset_bar <-ggplot(flow_subset, aes(month)) + geom_bar(aes(fill=extreme_type))
print(flow_subset_bar)
```

As can be seen from the bar graph, the majority of the maximum flow events occur in June, while most of the minimum flow events were observed in March. Moreover, there have not been any overlap in the months in which minimum and maximum events occur. In other words, there have not been a month, in which both a maximum and minimum flow was observed. 




### Exploring the characteristics of the extreme flow events

It may also be valuable to look in more details at the values of flow observed in each of the months. Hence, as a next step of the data analysis, I would like to present the relation between the month and the extreme flow in two graphs: One showing only maximum flow and another showing only minimum flow. This would allow to gain a better insight into the annual variations in the magnitude of each category of the extreme flow events.

To achieve this, two more subsets will be created - one for extreme_type "minimum" and one for extreme_type "maximum" - **using the filter() function**


```{r}
#Creting subset with only minimum extreme flow events
flow_subset_min <- filter(flow_subset, extreme_type=="minimum")

#Creating subset with only maximum extreme weather events:
flow_subset_max <- filter(flow_subset, extreme_type=="maximum")

```

#### Jitter plot

**Maximum flow:**

The following graph visualizes the extreme maximum flow events recorded in each of the months:

```{r}
flow_max_jitter <-ggplot(flow_subset_max, aes(month, flow)) + 
  geom_jitter(alpha = 0.4, width=0.2)
print(flow_max_jitter)
```

As can be seen from the graph, the majority and the highest magnitude of the maximum flow events occur in June. May and July have experienced fewer number of maximum flow events. Finally, August have only experienced one maximum flow event.

**Minimum flow:**

The following graph displays the extreme minimum flow events recorded in each of the months:

```{r}
flow_min_jitter <-ggplot(flow_subset_min, aes(month, flow)) + geom_jitter(alpha = 0.4, width=0.2)
print(flow_min_jitter)
```

As can be seen from the graph, minimum flow events occur in January, February, March, April, November and December. The lowest flow was observed in January, followed by February and March. 


#### Boxplot

Another way to visualize the values of flow observed in each of the months is by plotting boxplot.

```{r}
flow_min_boxplot <-ggplot(flow_subset_min, aes(month, flow))+ geom_boxplot(width = 0.2)+scale_y_log10()
print(flow_min_boxplot)
```

As can be seen from the boxplot, the lowest mean flow was observed in February, while the lowest value even recorded occurred in January.


A similar boxplot can be created for maximum flow:
```{r}
flow_max_boxplot <-ggplot(flow_subset_max, aes(month, flow))+ geom_boxplot(width = 0.2)+scale_y_log10()
print(flow_max_boxplot)
```

As can be seen from the boxplot, the highest mean of the maximum flow occurred in June. Moreover, June was the month when the highest flow was recorded,


### Research Questions
1. Have the maximum flow events increased in magnitude or frequency since 1909? 

2. When did the absolute maximum and minimum magnitude of the flow occur?

3. Does the annual distribution of the extreme flow events changed over time? For example, were most of the minimum/maximum that have occurred in past 10 years observed in a similar time than those in 1909-1919 or have the pattern shifted?

4. Have the number of data entries that are estimated or have significant gaps during the day (sym=E or sym=B, respectively) decreased over time?



### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.